---
title: "Reproducible Analysis on Activity Monitoring Data"
author: "Jiachang (Ernest) Xu"
date: "6/20/2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Section 1: Loading and preprocessing the data

First of all, we load the raw data, and take a look at the top 6 rows of the raw data.

```{r loading data, echo=TRUE}
## loading data from activity.csv
if(!exists("activity.raw")) {
    activity.raw <- read.csv("./activity.csv")
}
head(activity.raw)
```

We can immediately find some missing values in the "steps" column. Therefore, we have the need to process the raw data to make it analytic data. Steps of data cleanning include: (1) removing NA values in all three columns, (2) reformatting the "date" column to datetime objects, and (3) converting "interval" column into "interval.index" (a.k.a. the i-th 5-minute interval). After the data cleaning is done, we can take a quick look at the valid data frame.

```{r processing data, echo=TRUE}
## removing missing values
activity.valid <- activity.raw[!is.na(activity.raw$steps) & !is.na(activity.raw$date) & !is.na(activity.raw$interval), ]

## reformatting date object
activity.valid$date <- as.Date(activity.valid$date)

## converting interval into interval.index
colnames(activity.valid)[3] <- c("interval.index")
activity.valid$interval.index <- activity.valid$interval.index / 5

## glancing at activity.valid
head(activity.valid)
```

## Section 2: What is mean total number of steps taken per day?

We use the aggregate() function to find the total number of step taken per day.

```{r aggregating by date, echo=TRUE}
## finding sum of steps per day
activity.dailysteps <- aggregate(steps ~ date, activity.valid, sum)
```

We can plot a histogarm of total number of steps per day from the aggregated data from above.

```{r plotting histogram, echo=TRUE}
## plotting histogram of the total number of steps per day
hist(activity.dailysteps$steps, xlab = "Total number of steps per day", ylab = "Days", main = "Figure 1: Total Number of Steps per Day")
```

Then, we can easily calculate the avarage from the aggregated daily activity data.

```{r finding average daily steps, echo=TRUE}
mean(activity.dailysteps$steps)
```

## Section 3: What is the average daily activity pattern?

We use the aggregate() function to find the average number of step taken per 5-minute interval across the monitoring timeline from the valid data.

```{r aggregating by interval, echo=TRUE}
## find average steps per 5-minute interval
activity.pattern <- aggregate(steps ~ interval.index, activity.valid, mean)
```

We use ggplot2 package to plot a line graph of average number of step taken per 5-minute interval from the aggregated interval activity data.

```{r plotting daily activity, echo=TRUE}
## plotting daily activity pattern
require(ggplot2)
ggplot(activity.pattern, aes(interval.index, steps)) + geom_line() + xlab("The i-th 5-minute interval") + ylab("Average steps") + ggtitle("Figure 2: Average Daily Activity Pattern")
```

## Section 4: Imputing missing values

Please refer back to Section 1 for details. The histogram of total number of steps per day is already displayed in Section 2.

## Are there differences in activity patterns between weekdays and weekends?

First, we need to subset the valid data frame into weekdays and weekends data frame.

```{r subsetting weekdays and weekends, echo=TRUE}
## subsetting weekdays and weekends
activity.weekdays <- activity.valid[weekdays(activity.valid$date)!="Sunday" & weekdays(activity.valid$date)!="Saturday", ]
activity.weekends <- activity.valid[weekdays(activity.valid$date)=="Sunday" | weekdays(activity.valid$date)=="Saturday", ]
head(activity.weekdays)
head(activity.weekends)
```

We use the aggregate() function to find the average number of step taken per 5-minute interval for weekdays and weekends.

```{r aggregating by weekdays/weekends, echo=TRUE}
## aggregating by weekdays/weekends
activity.weekdays.pattern <- aggregate(steps ~ interval.index, activity.weekdays, mean)
activity.weekends.pattern <- aggregate(steps ~ interval.index, activity.weekends, mean)
```

Then, we combine two data frames into one.

```{r combining into one data frame, echo=TRUE}
activity.pattern.2 <- cbind(activity.weekdays.pattern, activity.weekends.pattern$steps)
colnames(activity.pattern.2)[2:3] <- c("weekdays.steps", "weekends.steps")
head(activity.pattern.2)
```

Finally, we use ggplot2 package to plot two lines of average number of step taken per 5-minute interval from the aggregated interval activity data, representing weekdays and weekends patterns.

```{r plotting average daily pattern by weekdays/weekends, echo=TRUE}
## plotting average daily pattern by weekdays/weekends
require(ggplot2)
ggplot(activity.pattern.2, aes(interval.index)) + 
    geom_line(aes(y = weekdays.steps, colour = "weekdays.steps")) + 
    geom_line(aes(y = weekends.steps, colour = "weekends.steps")) +
    xlab("The i-th 5-minute interval") + ylab("Average steps") + 
    ggtitle("Figure 3: Average Daily Activity Pattern by Weekdays/Weekends")
```

**Figure 3** shows that during weekdays, this person tends to have more activities during morning rush hours, and in the evening. However, this person has much less fluctuation during daytime on weekends.