---
title: "Regressional Analysis on Fuel Efficiency"
author: "Jiachang (Ernest) Xu"
date: "6/16/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary

This project will explore the relationship between miles per gallon (MPG) and other variables from the Motor Trends dataset. We are particularly interested in the following two questions:

 * Is an automatic or manual transmission better for MPG?
 * Quantify the MPG difference between automatic and manual transmissions
 
Our first guess is manual transmission is better for MPG than automatic transmission. The exploratory t.test verifies this guess.

```{r loading data, include=FALSE}
require(datasets)
require(plyr)
require(ggplot2)
require(GGally)
require(car)
data("mtcars")
attach(mtcars)
```

```{r splitting data, include=FALSE}
auto <- mtcars[which(am == 1), ]
manual <- mtcars[which(am == 0), ]
```

## Exploratory Hypothesis Testing

The following t.test and boxplots (Figure 1) show that the average MPG of manual transmission is significantly higher that that of automatic transmission. Therefore, based on the siginificant level of 5%, we can conclude that **manual transmission is better for MPG than automatic transmission**. Please refer to Appendix I about loading and splitting data.

```{r exploratory test, echo=TRUE}
t.test(manual$mpg, auto$mpg, alternative = "less")$p.value
```

```{r figure 1, echo=FALSE}
figure_1 <- ggplot(mtcars, aes(y=mpg,x=am)) + geom_boxplot(aes(fill=factor(am))) + labs(title = "Figure 1: Sample Distribution of MPG based on Transmission Types") + annotate("text", x = .5, y = 10, label = c("p = .00068")) + geom_jitter(width = 0.2) + annotate("text", x = 0, y = 16.85, label = c("mean = 17.15"))+ annotate("text", x = 1, y = 24.39, label = c("mean = 24.39")) + annotate("text", x = .5, y = 11, label = c("Corr = .6")) + annotate("text", x = .5, y = 12, label = c("Beta = 7.23"))

figure_1
```

## Model/Feature Selection

```{r figure 2, echo=FALSE}
pair_loess <- function(data, mapping, method="loess", ...){
       p <- ggplot(data = data, mapping = mapping) + 
              geom_point() + 
              geom_smooth(method=method, ...)
       p
}

g <- ggpairs(mtcars, lower = list(continuous=pair_loess))
```


## Conclusion

Our analysis shows that **manual transmission is more fuel eifficient than automatic transmission**.

## Appendix I: Loading and Splitting Data

 * Loading Data
 
```{r appendix loading data, eval=FALSE, echo=TRUE}
require(datasets)
require(plyr)
require(ggplot2)
require(GGally)
require(car)
data("mtcars")
attach(mtcars)
```

 * Splitting Data
 
```{r appendix splitting data, eval=FALSE, echo=TRUE}
auto <- mtcars[which(am == 1), ]
manual <- mtcars[which(am == 0), ]
```

## Appendix II: Figure 1 Plotting Code

```{r appendix figure 1, eval=FALSE, echo=TRUE}
figure_1 <- ggplot(mtcars, aes(y=mpg,x=am)) + geom_boxplot(aes(fill=factor(am))) + labs(title = "Figure 1: Sample Distribution of MPG based on Transmission Types") + annotate("text", x = .5, y = 10, label = c("p = .00068")) + geom_jitter(width = 0.2) + annotate("text", x = 0, y = 16.85, label = c("mean = 17.15"))+ annotate("text", x = 1, y = 24.39, label = c("mean = 24.39")) + annotate("text", x = .5, y = 11, label = c("Corr = .6")) + annotate("text", x = .5, y = 12, label = c("Beta = 7.23"))

figure_1
```

