---
title: "Regressive Analysis on Fuel Efficiency"
author: "Jiachang (Ernest) Xu"
date: "6/16/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary

This project will explore the relationship between miles per gallon (MPG) and other variables from the Motor Trends dataset. We are particularly interested in the following two questions:

 * Is an automatic or manual transmission better for MPG?
 * Quantify the MPG difference between automatic and manual transmissions
 
Our first guess is manual transmission is better for MPG than automatic transmission. The exploratory t.test verifies this guess. We try different regressors to find the optimal model based on the highest adjust R-squared with a significant p-value.

```{r loading data, include=FALSE}
require(datasets)
require(plyr)
require(ggplot2)
require(GGally)
require(car)

data(mtcars)
attach(mtcars)
```

```{r splitting data, include=FALSE}
auto <- mtcars[which(am == 1), ]
manual <- mtcars[which(am == 0), ]
```

## Exploratory Analysis

The following t.test and boxplots (Figure 1) show that the average MPG of manual transmission is significantly higher that that of automatic transmission. Therefore, based on the siginificant level of 5%, we can conclude that **manual transmission is better for MPG than automatic transmission**. Please refer to Appendix I about loading and splitting data.

```{r exploratory test, echo=TRUE}
t.test(manual$mpg, auto$mpg, alternative = "less")$p.value
```

```{r figure 1, echo=FALSE}
figure_1 <- ggplot(mtcars, aes(y=mpg,x=am)) + geom_boxplot(aes(fill=factor(am))) + labs(title = "Figure 1: Sample Distribution of MPG based on Transmission Types") + annotate("text", x = .5, y = 10, label = c("p = .00068")) + geom_jitter(width = 0.2) + annotate("text", x = 0, y = 16.85, label = c("mean = 17.15"))+ annotate("text", x = 1, y = 24.39, label = c("mean = 24.39")) + annotate("text", x = .5, y = 11, label = c("Corr = .6")) + annotate("text", x = .5, y = 12, label = c("Beta = 7.23"))

figure_1
```

## Model Selection

**Model 1** includes all variables, but has an insignificant p-value.
 
```{r model 1, include=TRUE, echo=TRUE}
model_1 <- lm(mpg~., mtcars)
model_1
```

```{r summary 1, echo=TRUE}
summary_1 <- append(summary(model_1)$coef[2,], c(summary(model_1)$adj.r.squared, summary(model_1)$r.squared))
names(summary_1)[5:6] <- c("Adj R^2", "R^2")
summary_1
```

**Model 2** includes am only, but only explains 34% of the variability.
 
```{r model 2, include=TRUE, echo=TRUE}
model_2 <- lm(mpg~am, mtcars)
model_2
```

```{r summary 2, echo=TRUE}
summary_2 <- append(summary(model_2)$coef[2,], c(summary(model_2)$adj.r.squared, summary(model_2)$r.squared))
names(summary_2)[5:6] <- c("Adj R^2", "R^2")
summary_2
```

**Model 3** uses algorithm to select optimal regressors. It has both higher variabiliry coverage and significant p-value.
 
```{r regressor selection, include=FALSE}
fitmem = {0}
memdex = {0}
    #Removing mpg and am from regressor list since they are necessary for our purposes
    regressors <- names(mtcars)[c(2:8,10:11)]
       
    #Generating a binary matrix for using each regressor or not in a grid matrix
    usematrix <- expand.grid(c(TRUE,FALSE), c(TRUE,FALSE),
                             c(TRUE,FALSE), c(TRUE,FALSE),c(TRUE,FALSE), c(TRUE,FALSE),
                             c(TRUE,FALSE), c(TRUE,FALSE)) 
       
    #Creates a list of all possible regressor cominations, there are 256
    Models <- apply(usematrix, 1, function(x) as.formula(paste(c("mpg ~ am", regressors[x]), collapse=" + ")))
    #Applies linear models to all of the possible functions and stores in list
    df <- lapply(Models,lm,data =mtcars)
       
    #Loops over the list storing the max value for adjusted R^2 and the corresponding index
    for (i in 1:256){
        if (summary(df[[i]])$adj.r.squared > fitmem) {
            fitmem <- summary(df[[i]])$adj.r.squared
                memdex <- i
            } 
        fitmem
        memdex
    }
```

```{r model 3, include=TRUE, echo=TRUE}
model_3 <- df[[memdex]]
model_3
```

```{r summary 3, echo=TRUE}
summary_3 <- append(summary(model_3)$coef[2,], c(summary(model_3)$adj.r.squared, summary(model_3)$r.squared))
names(summary_2)[5:6] <- c("Adj R^2", "R^2")
summary_3
```

## Conclusion

Our analysis shows that **manual transmission is more fuel eifficient than automatic transmission, by an average of 3.407 MPG higher holding all other variables constant**.

## Appendix I: Loading and Splitting Data

 * Loading Data
 
```{r appendix loading data, eval=FALSE, echo=TRUE}
require(datasets)
require(plyr)
require(ggplot2)
require(GGally)
require(car)
data("mtcars")
attach(mtcars)
```

 * Splitting Data
 
```{r appendix splitting data, eval=FALSE, echo=TRUE}
auto <- mtcars[which(am == 1), ]
manual <- mtcars[which(am == 0), ]
```

## Appendix II: Figure 1 Plotting Code

```{r appendix figure 1, eval=FALSE, echo=TRUE}
figure_1 <- ggplot(mtcars, aes(y=mpg,x=am)) + geom_boxplot(aes(fill=factor(am))) + labs(title = "Figure 1: Sample Distribution of MPG based on Transmission Types") + annotate("text", x = .5, y = 10, label = c("p = .00068")) + geom_jitter(width = 0.2) + annotate("text", x = 0, y = 16.85, label = c("mean = 17.15"))+ annotate("text", x = 1, y = 24.39, label = c("mean = 24.39")) + annotate("text", x = .5, y = 11, label = c("Corr = .6")) + annotate("text", x = .5, y = 12, label = c("Beta = 7.23"))

figure_1
```

## Appendix III: Model Selection

```{r appendix model 1, eval=FALSE, echo=TRUE}
model_1 <-lm(mpg~., mtcars)
model_1
```

```{r appendix summary 1, eval=FALSE, echo=TRUE}
summary_1 <- append(summary(model_1)$coef[2,], c(summary(model_1)$adj.r.squared, summary(model_1)$r.squared))
names(summary_1)[5:6] <- c("Adj R^2", "R^2")
summary_1
```

```{r appendix model 2, eval=FALSE, echo=TRUE}
model_2 <-lm(mpg~am, mtcars)
model_2
```

```{r appendix summary 2, eval=FALSE, echo=TRUE}
summary_2 <- append(summary(model_2)$coef[2,], c(summary(model_2)$adj.r.squared, summary(model_2)$r.squared))
names(summary_2)[5:6] <- c("Adj R^2", "R^2")
summary_2
```

```{r appendix model 3, eval=FALSE, echo=TRUE}
model_3 <- df[[memdex]]
model_3
```

```{r appendix summary 3, eval=FALSE, echo=TRUE}
summary_3 <- append(summary(model_3)$coef[2,], c(summary(model_3)$adj.r.squared, summary(model_3)$r.squared))
names(summary_2)[5:6] <- c("Adj R^2", "R^2")
summary_3
```