---
title: "Damage Analysis on Natural Disasters based on NOAA’s Storm Database"
author: "Jiachang (Ernest) Xu"
date: "6/21/2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Section 1: Synopsis

The objective of this project is to investigate U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database, to find which type of event has the greatest impact on population health, and poses the most severe economic consequences. We begin this analysis with loading data, and then extracting useful columns to form a clean dataset ready for further analysis. Then, we use aggregate() function to find the average fatalities, injuries, property damage, crop damage, and total damages by types of events, and extract the highest 5 types of events in each damage categories. Finally, we use the extracted top-5 data frames to plot barplots to communicate our findings.

## Section 2: Data Processing

Before data cleanning, we need to load the raw data set from StormData.csv. After data loading is finished, we take a look at the top 6 rows of the raw data set

```{r data loading, echo=TRUE}
## data loading
if (!exists("storm.raw")) {
    storm.raw <- read.csv("./data/StormData.csv")
}
head(storm.raw)
```

Since we are concerned with the relationship between types of events (**EVTYPE**) and population health (**FATALIITIES** & **INJURIES**) or economic consequences (**PROPDMG** & **CROPDMG**), we need to extract these 5 colmuns, and removing NA values for the purpose of data cleaning.

```{r data cleaning, echo=TRUE}
## extract the only 7 columns that we are interested in: 
## EVTYPE, FATALITIES, INJURIES, PROPDMG, CROPDMG
storm.interested <- storm.raw[, c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "CROPDMG")]

## remove any rows that have NA values in any of these 7 variables
storm.clean <- storm.interested[!is.na(storm.interested$EVTYPE) & 
                                   !is.na(storm.interested$FATALITIES) & 
                                   !is.na(storm.interested$INJURIES) & 
                                   !is.na(storm.interested$PROPDMG) & 
                                   !is.na(storm.interested$CROPDMG), ]

## convert all lowercase letters to uppercase
storm.clean$EVTYPE <- toupper(storm.clean$EVTYPE)

## take a look at the top 6 rows of the clean dataset
head(storm.clean)
```

## Section 3: Exploratory Analysis

 * **Types of Events vs. Population Health**

We can use aggregated() function to find the average fatalities and injuries by types of events, and store them into two new data frame. Combining these two new data frames grants us to reorder the data frame. Therefore, we can extract the top 5 types of events with highest average fatalities and injuries, separately and together. The extracted top-5 data frames are intended to be plotted in Section 4.

```{r analyze population health, echo=TRUE}
## find the average fatalities by types of events
storm.FATALITIES <- aggregate(FATALITIES ~ EVTYPE, storm.clean, mean)

## find the average injuries by types of events
storm.INJURIES <- aggregate(INJURIES ~ EVTYPE, storm.clean, mean)

## combine two data frame
storm.health <- cbind(storm.FATALITIES, storm.INJURIES$INJURIES)
colnames(storm.health)[3] <- "INJURIES"

## display the top 5 types of events with highest average fatalities and injuries, separately and together
storm.FATALITIES.top5 <- head(storm.FATALITIES[order(storm.FATALITIES$FATALITIES, decreasing = TRUE), ], 5)
storm.INJURIES.top5 <- head(storm.INJURIES[order(storm.INJURIES$INJURIES, decreasing = TRUE), ], 5)
storm.health.top5 <- head(storm.health[order(storm.health$FATALITIES, storm.health$INJURIES, decreasing = TRUE), ], 5)

storm.FATALITIES.top5
storm.INJURIES.top5
storm.health.top5
```
 
 * **Types of Events vs. Economic Consequences**

We can use aggregated() function to find the average property and crop damages by types of events, and store them into two new data frame. Combining these two new data frames grants us to reorder the data frame. Therefore, we can extract the top 5 types of events with highest property, crop, and total damages on average. The extracted top-5 data frames are intended to be plotted in Section 4.

```{r analyze economic consequences, echo=TRUE}
## find the average property damage by types of events
storm.PROPDMG <- aggregate(PROPDMG ~ EVTYPE, storm.clean, mean)

## find the average crop damage by types of events
storm.CROPDMG <- aggregate(CROPDMG ~ EVTYPE, storm.clean, mean)

## combine two data frame
storm.economic <- cbind(storm.PROPDMG, storm.CROPDMG$CROPDMG)
colnames(storm.economic)[3] <- "CROPDMG"
storm.economic$TOTALDMG <- storm.economic$PROPDMG + storm.economic$CROPDMG

## extract the top 5 type of events with highest average property and crop damages, seperately and together
storm.PROPDMG.top5 <- head(storm.PROPDMG[order(storm.PROPDMG$PROPDMG, decreasing = TRUE), ], 5)
storm.CROPDMG.top5 <- head(storm.CROPDMG[order(storm.CROPDMG$CROPDMG, decreasing = TRUE), ], 5)
storm.economic.top5 <- head(storm.economic[order(storm.economic$TOTALDMG, decreasing = TRUE), ], 5)

## display the top 5 type of events with highest average property and crop damages, seperately and together
storm.PROPDMG.top5
storm.CROPDMG.top5
storm.economic.top5
```

## Section 4: Results

 * **Types of Events vs. Population Health**
 
The data frames that store the top 5 types of events with highest average fatalities and injuries are already extracted in Section 3. We can plot them, using barplot in descending order.
 
```{r plot population health, echo=TRUE}
require(ggplot2)
require(gridExtra)

## reorder
storm.FATALITIES.top5 <- transform(storm.FATALITIES.top5, EVTYPE = reorder(EVTYPE, -FATALITIES))
storm.INJURIES.top5 <- transform(storm.INJURIES.top5, EVTYPE = reorder(EVTYPE, -INJURIES))

plot1.1 <- ggplot(data = storm.FATALITIES.top5, aes(x = EVTYPE, y = FATALITIES)) + 
    geom_bar(stat = "identity", fill = "firebrick") +
    geom_text(aes(label = as.integer(FATALITIES)), vjust = 1.6, color = "black", size = 3.5) +
    xlab("Types of Events") + ylab("Fatalities") + 
    ggtitle("Figure 1.1: Top 5 Types of Events with Highest Fatalities")

plot1.2 <- ggplot(data = storm.INJURIES.top5, aes(x = EVTYPE, y = INJURIES)) + 
    geom_bar(stat = "identity", fill = "orange3") +
    geom_text(aes(label = as.integer(INJURIES)), vjust = 1.6, color = "black", size = 3.5) +
    xlab("Types of Events") + ylab("Injuries") + 
    ggtitle("Figure 1.2: Top 5 Types of Events with Highest Injuries")

grid.arrange(plot1.1, plot1.2, ncol=1)
```

Figure 1.1 shows that **TORNADOES, TSTM WIND, HAIL** has the highest average fatalities across all types of events; Figure 1.2 shows that **TROPICAL STORM GORDON** has the highest average injuries across all types of events.

 * **Types of Events vs. Economic Consequences**

The data frames that store the top 5 types of events with highest property, crop, and total damages on average are already extracted in Section 3. We can plot them, using barplot in descending order.

```{r plot economic consequences, echo=TRUE}
require(ggplot2)
require(gridExtra)

## reorder
storm.PROPDMG.top5 <- transform(storm.PROPDMG.top5, EVTYPE = reorder(EVTYPE, -PROPDMG))
storm.CROPDMG.top5 <- transform(storm.CROPDMG.top5, EVTYPE = reorder(EVTYPE, -CROPDMG))
storm.economic.top5 <- transform(storm.economic.top5, EVTYPE = reorder(EVTYPE, -TOTALDMG))

plot2.1 <- ggplot(data = storm.PROPDMG.top5, aes(x = EVTYPE, y = PROPDMG)) + 
    geom_bar(stat = "identity", fill = "olivedrab") +
    geom_text(aes(label = as.integer(PROPDMG)), vjust = 1.6, color = "black", size = 3.5) +
    xlab("Types of Events") + ylab("Property Damage") + 
    ggtitle("Figure 2.1: Top 5 Types of Events with Highest Property Damage")

plot2.2 <- ggplot(data = storm.CROPDMG.top5, aes(x = EVTYPE, y = CROPDMG)) + 
    geom_bar(stat = "identity", fill = "steelblue") +
    geom_text(aes(label = as.integer(CROPDMG)), vjust = 1.6, color = "black", size = 3.5) +
    xlab("Types of Events") + ylab("Crop Damage") + 
    ggtitle("Figure 2.2: Top 5 Types of Events with Highest Crop Damages")

plot2.3 <- ggplot(data = storm.economic.top5, aes(x = EVTYPE, y = TOTALDMG)) + 
    geom_bar(stat = "identity", fill = "blueviolet") +
    geom_text(aes(label = as.integer(TOTALDMG)), vjust = 1.6, color = "black", size = 3.5) +
    xlab("Types of Events") + ylab("Total Damages") + 
    ggtitle("Figure 2.3: Top 5 Types of Events with Highest Total Damages")

grid.arrange(plot2.1, plot2.2, plot2.3, nrow=3)
```

Figure 2.1 shows that **COASTAL EROSION** has the highest average property damage across all types of events; Figure 2.2 shows that **DUST STORM/HIGH WINDS** has the highest average crop damage across all type of events; Figure 2.3 shows that **TROPICAL STORM GORDON** has the highest total damages on average.